### 2017-12-20 chapters/introduction/summary.tex
Add latest version of summary section

In this paper we develop and test the pipeline for constructing a
patient specific mechanical simulation of a patient's heart, based on
clinical measurements and adjoint-based data assimilation
techniques.

As a model case we consider one patient, diagnosed with
left bundle branch block and selected for cardiac resynchronization
therapy (CRT). Prior to the CRT implantation the patient had 4D
echocardiography taken for which the LV geometry, LV volumes and LV
regional strains throughout the cardiac cycle are measured. During
implantation of the CRT device the LV pressure where also measured
invasively. The LV pressure measurements are used as boundary condition at the
endocardium, while the LV volume and LV regional strain is incorporated
into a cost functional that we want to minimize.

The pipeline is divided into two phases, a passive phase where we
estimate the linear isotropic parameters $a$ in
\eqref{eq:holzapel_trans} as a global material parameter using the
measurement points belonging atrial systole, and an active phase where we estimate the
a spatially varying contraction parameter ($\gamma$ in
\eqref{eq:intro_active_strain_Fa_gjerald}) at each measurement point
with active contraction. During the passive phase we only fit the
volumes, while during the active phase 51 strain measurements in the radial,
longitudinal and circumferential direction in each AHA segment (Figure
\ref{fig:echopac_output}) are used in the optimization.

The results show an excellent fit with measured strain an volume, with
a average relative error in the volume and strain of less than 0.4 \% and 3 \%
respectively using a contraction parameter with one degree of freedom
for each vertex in the geometry (2661 parameters in total). Parameters
at lower spatial resolution are also tested to show the necessity of
high spatial resolution to fit the measured strain data.
A synthetic test is also performed in order to show that that method
is able to recapitulate the generated data, also with noise added to
the data. Moreover, a sensitivity analysis to different parameters are
presented in the appendix.

@@

In this paper we apply the method developed in the previous paper to a
cohort of patients and estimate indices of cardiac contractility. More
specifically, a group of seven patients diagnosed with left bundle branch block and
selected for cardiac resynchronization therapy (CRT) and a group of
seven healthy control subjects where included in the study.

@@

The optimized active strain parameter in
\eqref{eq:intro_active_strain_Fa_gjerald} as well as the optimized
active stress parameter in \eqref{eq:intro_active_stress} are averaged
over the ventricle and compared between the two groups. The healthy
group showed a significant increase in both of these
parameters. Furthermore, estimation of end-systolic elastance by
perturbation of the pressure at the end-systolic state while fixing
the remaining quantities were made. This estimate of end-systolic
elastance where also significantly higher in the healthy control
group.

\subsection{Paper 3: Personalized bi-ventricular mechanical analysis
  using gradient-based optimization..}

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.8\textwidth]{chapters/introduction/figures/paper3}
\caption{Summary figure for paper 3}
\label{fig:paper3}
\end{figure}

@@

In this work we estimate the linear isotopic material parameter in
\eqref{eq:holzapel_trans}, spatially resolved on the LV and RV, by
minimizing the error in end-diastolic LV and RV cavity volumes. We also
apply a simplified algorithm for estimating in unloaded geometry, in
order to overcome issues related to estimating an unloaded
BiV-geometry, which is not well-posed. The amount of active
contraction is estimated during the active phase, by minimizing
simulated and measured volumes and circumferential strain. The active
control parameter, which are $\gamma$ in
\eqref{eq:intro_active_strain_Fa_gjerald}  and $T_a$ in
\eqref{eq:intro_active_stress} for the active strain and active stress
formulation respectively, are spatially resolved on the
LV free wall (LVFW), the RV free wall (RVFW) and the septum. With so
few control parameters, issues related to conflicting objectives, i.e
that is is not possible to minimize both the strain and volume, is
evident. However, with a low dimensional parameter space, questions
regarding identifiability and uniqueness of the estimated parameters
are easier to answer. In this study we also perform a validation of
the model, by comparing simulated and measured longitudinal strain
which is not used in the optimization.

@@

In this paper we investigate if it is possible to assess regional
myocardial work using data assimilation and finite element modeling.
The amount of work performed by the left ventricle equals to the
amount to work needed to eject blood against the intraventricular
pressure. Effectively, a global measure of work is therefore the same
as the area of the pressure volume (PV) loop which is most commonly refered
to as stroke work.  However, stroke work is a global measure which
does not reflect whether some regions in the ventricle work harder than
others to compensate for regional dysfunction, in order to maintain the
stroke work needed to meet the body's demand. Therefore a regional
measure of work would be helpful in assessment of patients with
regional dysfunction, and preferable we would have a regional meausure
that summed up to the actual stroke work. The area of the
pressure-strain loops, where strain here refers to as regional
longitudinal strain, has been proposed as a meausure regional
myocardial work, and indices based on these work calculations have
shown to be a better predictor response to cardiac resynchronization
therapy, a treatment given to patients with regional dysfunction.

@@

\subsection{Talks}
\begin{itemize}
  \item Henrik Finsberg, Gabriel Balaban, Joakim Sundnes, Hans
    Henrik Odland, Marie Rognes, and Samuel T. Wall. ``Patient
    Constrained Ventricular Stress Mapping'',
    Conference Presentation at MALT 2015,  Lugano, Switzerland (2015).
  \item Henrik Finsberg, Gabriel Balaban, Joakim Sundnes, Marie
    Rognes, and Samuel T. Wall. ``Personalization of a Cardiac
    Computational Model using Clinical Measurements'', Conference
    Presentation at 28th Nordic Seminar on Computational
    Mechanics. Vol. 28. Tallin, Estonia, (2015).
  \item Henrik Finsberg, Gabriel Balaban, Joakim Sundnes, Marie
    Rognes, and Samuel T. Wall. ``Optimization of a Spatially Varying
    Cardiac Contraction parameter using the Adjoint Method'',
    Conference Presentation at FEniCS 16, Oslo, Norway,(2016).
  \item Finsberg, Henrik N., Gabriel Balaban, Joakim Sundnes, Hans
    Henrik Odland, Marie Rognes, and Samuel T. Wall. ``Personalized
    Cardiac Mechanical Model using a High Resolution Contraction Field
    '',  Conference Presentation at VPH16 Translating VPH to the
    Clinic,  Amsterdam, Netherlands (2016).
\end{itemize}

@@

Although we have shown in this thesis that the techniques developed in
this thesis are powerful, and opens up new possibilities in terms of
patient specific mechanical simulations, many questions still has to
be answered before we can fully embrace the output of such simulations.

First of all, is should be clear that \emph{the
  quality of biomarkers you can extract from a data-driven
  model cannot be any better than the data used as input to the
  model}. A typical saying is that garbage in $=$ garbage out,
meaning that if the data you use to constrain the model is noisy, then
you will also fit this noise if you allow for enough degree of
freedom. Regularization techniques (Section
\ref{sec:intro_regularization}) provides a way to attack this problem,
but it is not clear what is the best approach. If the noise in the
data is normally distributed with zero expected value, then adding
more data to the cost functional will also have a regularizing effect.
Hence, as the data assimilation techniques developed in this thesis
can take into account large amount of data, as much data as possible
should be provided to the assimilator.

A rule of thump is that \emph{the spatial resolution of the parameters should be
  reflected in the spatial resolution of the observations}. This means
that if one is trying to fit data that are spatially resolved at some level,
then choosing parameters that are resolved at a finer level should be
done with caution. Regarding both paper 1, 2 and 4 we see that the
spatial resolution chosen was at a much finer level than the input
data. In this case, regularization techniques was used to restrict the
parameter space, by assuming that the solution you are seeking has
some particular properties, e.g smoothness.

Regarding the mechanical modeling of the heart, 
\emph{choosing appropriate boundary conditions} that reflect the reality has
been an issue during the work of this thesis, and several different
choices has been made. Moreover, \emph{accounting for the orthotropic as
well as the visco-elastic behavior of the myocardium} is something that should
be investigated in future studies. Accounting for an orthotropic
behavior would acquire more parameters to be estimated, which would
thus require more input data.

In this thesis we have also not fully
explored the \emph{spatial resolution of the material parameters}, and it
should be investigated whether it is possible to relate locally estimated
tissue stiffness to e.g myocardial infarction.

% The choice of active model for the myocardium is another topic that
% should be addressed in more detail in future studies. In the author's
% opinion, the output of the two fundamentally different approaches may
% vary a lot. In particular, it is evident that the active strain
% formulation do a better job in fitting strain data. One
% hypothesis is that the amount of transverse active stresses should be
% adjusted to each individual. These transverse active stresses are
% naturally embedded in the active strain approach via a volume
% preserving active deformation.

Finally, when estimating high dimensional parameters, a natural question
concerning uniqueness of these estimates arises. It is obvious that if one allows for
enough degree of freedom in the parameter space, then it is possible
to fit almost any type of data. Therefore \emph{more work on ensuring
identifiability} of these estimates (Section
\ref{sec:intro_identifiability}) should be done. If the output of such
model should have any clinical utility, then uniqueness of the
estimated parameters is absolutely pivotal. Moreover, \emph{validation} of
these model is what matter most in terms of translating such computational
models into the clinic.
% Here we start by listing a couple of statements
% which should be taken into considerations.

>>>>

In this paper we develop and test a pipeline for constructing a
patient-specific mechanical simulation of a patient's heart, based on
clinical measurements and adjoint-based data assimilation
techniques.

As a model case we consider one patient, diagnosed with
left bundle branch block and selected for cardiac resynchronization
therapy (CRT). Prior to the CRT implantation the patient had 4D
echocardiography taken for which the LV geometry, LV volumes and LV
regional strains throughout the cardiac cycle are measured. During
implantation of the CRT device the LV pressure were also measured
invasively. The LV pressure measurements were used as boundary condition at the
endocardium, while the LV volume and LV regional strain were incorporated
into a cost functional that we wanted to minimize.

The pipeline is divided into two phases, a passive phase where we
estimate the linear isotropic parameters $a$ in
\eqref{eq:holzapel_trans} as a global material parameter using the
measurement points belonging atrial systole, and an active phase where we estimate
a spatially varying contraction parameter ($\gamma$ in
\eqref{eq:intro_active_strain_Fa_gjerald}) at each measurement point
with active contraction. During the passive phase we only fit the
volumes, while during the active phase a total of 51 strain measurements in the radial,
longitudinal and circumferential direction in each AHA segment (Figure
\ref{fig:echopac_output}) were used additionally in the optimization.

The results show an excellent fit with measured strain an volume, with
a average relative error in the volume and strain of less than 0.4 \% and 3 \%
respectively using a contraction parameter with one degree of freedom
for each vertex in the geometry (2661 parameters in total). Parameters
at lower spatial resolution are also tested to show the necessity of
high spatial resolution to fit the measured strain data.
A synthetic test is also performed in order to show that the method
is able to recapitulate the generated data, also with noise added to
the data. Moreover, a sensitivity analysis to different parameters are
presented in the appendix.

@@

In this paper we apply the method developed in the previous paper to a
cohort of patients and estimate indices of cardiac contractility. More
specifically, a group of seven patients diagnosed with left bundle branch block and
selected for cardiac resynchronization therapy (CRT), and a group of
seven healthy control subjects were included in the study.

@@

The optimized active strain parameter in
\eqref{eq:intro_active_strain_Fa_gjerald} as well as the optimized
active stress parameter in \eqref{eq:intro_active_stress} are averaged
over the ventricle and compared between the two groups. The healthy
group showed a significant increase in both of these
parameters. Furthermore, an estimation of the end-systolic elastance by
perturbation of the model at the end-systolic state, while fixing
the remaining quantities, was also conducted. This estimate of end-systolic
elastance were also significantly higher in the healthy control
group.

\subsection{Paper 3: Efficient estimation of personalized biventricular mechanical
  function employing gradient-based optimizatio}

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.75\textwidth]{chapters/introduction/figures/paper3}
\caption{Summary figure for paper 3}
\label{fig:paper3}
\end{figure}

@@

In this work we estimate the linear isotopic material parameter in
\eqref{eq:holzapel_trans}, spatially resolved on the LV and RV, by
minimizing the error in end-diastolic LV and RV cavity volumes.
We use a simplified algorithm to estimate the unloaded geometry, since
the problem of estimating an unloaded BiV-geometry is not well-posed
because buckling of the RV free wall might occur. The amount of active
contraction is estimated during the active phase, by minimizing
simulated and measured volumes and circumferential strain. The active
control parameter, which are $\gamma$ in
\eqref{eq:intro_active_strain_Fa_gjerald}  and $T_a$ in
\eqref{eq:intro_active_stress} for the active strain and active stress
formulation respectively, are spatially resolved on the
LV free wall (LVFW), the RV free wall (RVFW) and the septum. With so
few control parameters, issues related to conflicting objectives, i.e
that is is not possible to minimize both the strain and volume, is
evident. However, with a low dimensional parameter space, questions
regarding identifiability and uniqueness of the estimated parameters
are easier to answer. In this study we also perform a validation of
the model, by comparing simulated and measured longitudinal strain
which is not used in the optimization.

@@

In this paper we investigate if it is possible to assess regional
myocardial work using data assimilation and finite element modeling.
The amount of work performed by the left ventricle equals to the
amount to work needed to eject blood against the intraventricular
pressure. Effectively, a global measure of work is therefore the same
as the area of the pressure volume (PV) loop which is most commonly refered
to as stroke work.  However, stroke work is a global measure which
does not reflect whether some regions in the ventricle work harder than
others to compensate for regional dysfunction, in order to maintain the
stroke work needed to meet the body's demand. Therefore a regional
measure of work would be helpful in assessment of patients with
regional dysfunction, and preferable we would have a regional meausure
that summed up to the actual stroke work. The area of the
pressure-strain loops, where strain here refers to as regional
longitudinal strain, has been proposed as a meausure regional
myocardial work, and indices based on these work calculations have
shown to be a good predictor of response to cardiac resynchronization
therapy, a treatment given to patients with regional dysfunction.

@@

\subsection{Talks}
\begin{itemize}
  \item Henrik Finsberg, Gabriel Balaban, Joakim Sundnes, Hans
    Henrik Odland, Marie Rognes, and Samuel T. Wall. ``Patient
    Constrained Ventricular Stress Mapping'',
    Conference Presentation at MALT 2015,  Lugano, Switzerland (2015).
  \item Henrik Finsberg, Gabriel Balaban, Joakim Sundnes, Marie
    Rognes, and Samuel T. Wall. ``Personalization of a Cardiac
    Computational Model using Clinical Measurements'', Conference
    Presentation at 28th Nordic Seminar on Computational
    Mechanics. Vol. 28. Tallin, Estonia, (2015).
  \item Henrik Finsberg, Gabriel Balaban, Joakim Sundnes, Marie
    Rognes, and Samuel T. Wall. ``Optimization of a Spatially Varying
    Cardiac Contraction parameter using the Adjoint Method'',
    Conference Presentation at FEniCS 16, Oslo, Norway,(2016).
  \item Henrik Finsberg, Gabriel Balaban, Joakim Sundnes, Hans
    Henrik Odland, Marie Rognes, and Samuel T. Wall. ``Personalized
    Cardiac Mechanical Model using a High Resolution Contraction Field
    '',  Conference Presentation at VPH16 Translating VPH to the
    Clinic,  Amsterdam, Netherlands (2016).
  \item Henrik Finsberg, John Aalen,Camilla K.
    Larsen, Espen Remme, Joakim Sundnes, Otto A. Smiseth, and 
    Samuel T. Wall. ``Assessment of regional myocardial work via through
    adjoint-based data assimilation. '',  Conference Presentation at
    the International Conference on Computational Science and
    Enfineering, Oslo, Norway (2016).
\end{itemize}

@@

Although we have shown that the techniques developed in
this thesis are powerful, and opens up new possibilities in terms of
patient-specific mechanical simulations, many questions still have to
be answered before we can fully embrace the output of such simulations.

First of all, is should be clear that \emph{the
  quality of biomarkers you can extract from a data-driven
  model cannot be any better than the data used as input to the
  model}. A typical saying is that garbage in $=$ garbage out,
meaning that if the data you use to constrain the model is noisy, then
you will also fit this noise if you allow for enough degree of
freedom. Regularization techniques (Section
\ref{sec:intro_regularization}) provides a way to attack this problem,
but it is not clear what is the best approach.

A rule of thump is that \emph{the spatial resolution of the parameters should be
  reflected in the spatial resolution of the observations}. This means
that if one is trying to fit data that are spatially resolved at some level,
then choosing parameters that are resolved at a finer level should be
done with caution. Regarding both paper 1, 2 and 4 we see that the
spatial resolution chosen was at a much finer level than the input
data. In this case, regularization techniques were used to restrict the
parameter space, by assuming that the solution we seek had specific
properties, e.g smoothness.

Regarding the mechanical modeling of the heart, 
\emph{choosing appropriate boundary conditions} that reflect the reality has
been an issue during the work of this thesis, and several different
choices has been made. Moreover, \emph{accounting for the orthotropic as
well as the visco-elastic behavior of the myocardium} is something that should
be investigated in future studies. Accounting for an orthotropic
behavior would acquire more parameters to be estimated, which would
thus require more input data.
In this thesis we have also not fully
explored the \emph{spatial resolution of the material parameters}, and it
should be investigated whether it is possible to relate locally estimated
tissue stiffness to e.g myocardial infarction.
% The choice of active model for the myocardium is another topic that
% should be addressed in more detail in future studies. In the author's
% opinion, the output of the two fundamentally different approaches may
% vary a lot. In particular, it is evident that the active strain
% formulation do a better job in fitting strain data. One
% hypothesis is that the amount of transverse active stresses should be
% adjusted to each individual. These transverse active stresses are
% naturally embedded in the active strain approach via a volume
% preserving active deformation.

Finally, when estimating high dimensional parameters, a natural question
concerning uniqueness of these estimates arises. It is obvious that if one allows for
enough degree of freedom in the parameter space, then it is possible
to fit almost any type of data. Therefore \emph{more work on ensuring
identifiability} of these estimates should be done. If the output of such
models should have any clinical utility, then uniqueness of the
estimated parameters is absolutely pivotal. Moreover, \emph{validation} of
these model is what matter most in terms of translating such computational
models into the clinic.
% Here we start by listing a couple of statements
% which should be taken into considerations.
